Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-07 paper.tex
minor changes

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. 
Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

The “Multiple Merger” model has better log-likelihood than the “Continuous
Migration” model (-102,600 vs. -116,000, respectively). The two models differ
in the timing of divergence and the relative Ne of the two stems
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

>>>>

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
\sgcomment{IF we were not afraid of the racists, we could write something like:
"Despite the genetic similarity between these populations, 4 to 10\% of random
genetic differences among [modern human populations] can be attributed to this
early genetic drift.}
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

The “Multiple Merger” model has better log-likelihood than the “Continuous
Migration” model (-102,600 vs. -115,500, respectively). The two models differ
in the timing of divergence and the relative Ne of the two stems
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

### 2022-01-07 supplement.tex
minor changes

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
        $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
        amount of drift between sampled populations $A$ and $B$ that can be
        attributed to drift between populations $X$ and $Y$, sampled in the past
        (see Section~\ref{sec:f4}). \aprcomment{\dots}
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-continuous-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the continuous-migration model.}
    }
    \label{fig:supp-f4s-continuous-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-merger-without-stem-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the merger-without-stem-migration model.}
    }
    \label{fig:supp-f4s-merger-without-stem-migration}
\end{figure}

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
        $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
        amount of drift between sampled populations $A$ and $B$ that aligns with
        drift between populations $X$ and $Y$, sampled in the past
        (see Section~\ref{sec:f4}). In the single-origin model, nearly all
        differentiation between the Nama, Mende, and Gumuz populations arises
        after the divergence events, so normalized $f_4$ decays to zero by the
        time of those divergences.
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-continuous-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the continuous-migration model.}
        Despite population structure inferred to have extended up to 1Ma
        into the past, differentiation between modern African populations
        only traces back to differentiation between ancestral populations
        $\sim$100-200ka, indicating that while stem populations were structured
        (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
        ancient structure does not contribute to differention between modern
        populations.
    }
    \label{fig:supp-f4s-continuous-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-merger-without-stem-migration.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the merger-without-stem-migration model.}
        Unlike the continuous-migration model, 
    }
    \label{fig:supp-f4s-merger-without-stem-migration}
\end{figure}
